Reproducible Research: Peer Assessment 1
========================================================

## Pull the data file from source URL
``` {r, echo=TRUE}
setInternet2(use = TRUE)
sourceURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(sourceURL, "ActivityMonitoring.zip", mode="wb")
```


## Loading and preprocessing the data
The file is unzipped, and loaded into a data frame.  The date column is converted to a 'Date' data type.
```{r, echo=TRUE, cache=TRUE}
unzip("ActivityMonitoring.zip")
inDF <- read.csv("activity.csv", header=TRUE, colClasses=c("numeric","character",  
                                                           "numeric"))
inDF[,2] <- as.Date(inDF[,2], format="%Y-%m-%d")
```

## What is mean total number of steps taken per day?
```{r, echo=TRUE}
agDF <- aggregate(inDF$steps, list(inDF$date), sum, na.rm=TRUE )
hist(agDF[,2], col="grey", main="Total Number of Steps Taken Each Day", xlab="Number of Steps", ylab="Number of Days")
agMean <- mean(agDF[,2])
agMedian <- median(agDF[,2])
```

The mean and median values of the number of steps taken each day are as follows:
- Mean: `r agMean`
- Median: `r agMedian`


## What is the average daily activity pattern?
```{r, echo=TRUE}
dailyAvgDF <- aggregate(inDF$steps, list(inDF$interval), mean, na.rm=TRUE)
colnames(dailyAvgDF) <- c("interval", "avgSteps")
plot(dailyAvgDF[,1], dailyAvgDF[,2],type="l", main="Average Daily Activity Pattern", xlab="Intervals", ylab="Avg. # of Steps Taken")
maxVals <- subset(dailyAvgDF, avgSteps==max(avgSteps))
```

The interval that contains the maximum number of steps, on average across all the days in the dataset is `r maxVals[1,1]` and the maximum number of steps pertaining to this interval is `r maxVals[1,2]`

## Imputing missing values
The number of rows with NA is calculated as follows:
```{r, echo=TRUE}
numNA <- sum(is.na(inDF[,1]))
```
The number of rows with missing values is `r numNA`

The missing values are filled in using the mean value across all dates for the corresponding time interval.
```{r, echo=TRUE}
newDF <- inDF
numIterations <- nrow(inDF)
for (i in 1:numIterations)
{
  if(is.na(newDF[i,1]))
  {
    tsInt <- inDF[i,3]
    tempDF <- subset(dailyAvgDF, interval == tsInt)
    newDF[i,1] <- tempDF[1,2]
  }
}
agDF2 <- aggregate(newDF$steps, list(inDF$date), sum, na.rm=TRUE )
hist(agDF2[,2], col="grey", main="Total Number of Steps Taken Each Day", xlab="Number of Steps", ylab="Number of Days")
agMean2 <- mean(agDF2[,2])
agMedian2 <- median(agDF2[,2])
```

The mean and median values of the number of steps taken each day, after filling in missing values are as follows:
- Mean: `r agMean2`
- Median: `r agMedian2`


## Are there differences in activity patterns between weekdays and weekends?
``` {r, echo=TRUE}
#add column for category variable weekend/weekday
newDF$dayname <- weekdays(newDF$date)
newDF$day <- ifelse(newDF$dayname==c("Saturday", "Sunday"), "weekend", "weekday")

#split the dataframe by weekday/weekend and aggregate by intervals
weekendDF <- subset(newDF, day=="weekend")
weekendAvgDF <- aggregate(weekendDF$steps, list(weekendDF$interval), mean, 
                          na.rm=TRUE)

weekdayDF <- subset(newDF, day=="weekday")
weekdayAvgDF <- aggregate(weekdayDF$steps, list(weekdayDF$interval), mean, 
                          na.rm=TRUE)

#plot the two graphs
par(mfrow = c(2,1))
plot(weekendAvgDF[,1], weekendAvgDF[,2],type="l", main="Weekend Activity Pattern",
     xlab="Interval", ylab="Avg. # of Steps")
plot(weekdayAvgDF[,1], weekdayAvgDF[,2],type="l", main="Weekday Activity Pattern",
     xlab="Interval", ylab="Avg. # of Steps")
```



